---
title: "Exploration of Central Limit Theorem"
author: "Ben Kesseler"
date: "June 19, 2016"
output: pdf_document
geometry: margin=1cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

I will create 1000 simulations of a data set with 40 data points from an
exponential distribution. In addition, I will explore the mean and variance of
the means of those 1000 data sets, in comparison with their theoretical values.

# Simulations

First I need to create the data for my simulation. I'll create a matrix with
1000 rows, 40 columns, where each cell represents a pull from an exponential
distribution.

```{r libraries}
library(ggplot2)
```

```{r simulation}
lambda <- 0.2 # Setting the rate for the exponential distribution
n <- 40 # Setting the number of samples to take, per average
simulations <- 1000 # Setting the number of total simulations

# Set Seed for Reproducibility
set.seed(314152)

# Creating the overall simulations
matrix_simulations <- matrix(data = rexp(n*simulations, rate = lambda),
                             nrow = simulations)

# Creating the averages
simulation_means <- data.frame(means = apply(matrix_simulations, 1, mean))
```

# Sample Mean versus Theoretical Mean

The theoretical mean should be 1/lambda = `r 1/lambda`. I'll now plot the
distribution, including the sample mean and the theoretical mean, along with
the difference.

```{r meancomparison, fig.width = 10, fig.cap = "The plot shows the probability distribution of my simulations, along with the theoretical mean."}
annotate1 <- paste("Black Line = Sampled Mean =",
                   round(mean(simulation_means$means), 3))
annotate2 <- paste("Red Line = Theoretical Mean =", 1/lambda)
annotate3 <- paste("Difference from Theory =",
                   round(mean(simulation_means$means) - 1/lambda, 3))

ggplot(data = simulation_means, aes(x = means)) +
        geom_histogram(aes(y = ..density..),
                       binwidth = 0.1,
                       col = "black",
                       fill = "white") + 
        geom_density(col = "blue",
                     size = 2,
                     fill = "blue",
                     alpha = 0.3) + 
        theme_bw() + 
        ggtitle("Distribution of 1000 Simulations of the Mean of 40 Exponential Data Points") +
        geom_vline(xintercept = mean(simulation_means$means),
                   size = 1,
                   col = "black") +
        geom_vline(xintercept = 1/lambda,
                   size = 1,
                   col = "red") +
        annotate("text", x = 6, y = 0.5, label = annotate1, hjust = 0) +
        annotate("text", x = 6, y = 0.45, label = annotate2, hjust = 0) +
        annotate("text", x = 6, y = 0.4, label = annotate3, hjust = 0)
```

So, the mean I calculate is close to the theoretical mean, with a difference of
only `r mean(simulation_means$means) - 1/lambda`.

# Sample Variance versus Theoretical Variance

The theoretical variance is 1/lambda<sup>2</sup> = `r 1/(lambda^2)`. I'll now
calculate the sample variance, and the difference from the theoretical variance.

```{r variancecomparison}
# Theoretical Variance
theoretical_variance <- (1/(lambda*sqrt(n)))^2
theoretical_variance

# Sample Variance
sample_variance <- var(simulation_means$means)
sample_variance

# Difference in Variance
variance_difference <- sample_variance - theoretical_variance
variance_difference
```

So the sample variance is quite close to the theoretical variance, which is as
expected.

# Distribution

Now I will plot the distribution of my samples, along with the normal 
distribution I would expect to see, which would be a normal distribution with
the mean equal to 1/lambda = `r 1/lambda` and the standard deviation  equal to
1/(lambda * sqrt(n)) = `r sqrt(theoretical_variance)`.

```{r distributioncomparison, fig.width = 10, fig.cap = "Comparison of Sampled Distribution and Expected Theoretical Normal Distribution, 1000 simulations"}
annotate4 <- paste("Blue Line = Sampled Distribution")
annotate5 <- paste("Red Line = Theoretical Distribution")

ggplot(data = simulation_means, aes(x = means)) +
        geom_histogram(aes(y = ..density..),
                       binwidth = 0.1,
                       col = "black",
                       fill = "white") + 
        geom_density(col = "blue",
                     size = 2,
                     fill = "blue",
                     alpha = 0.3) + 
        stat_function(fun = dnorm,
                      args = list(mean = 1/lambda, sd = sqrt(theoretical_variance)),
                      colour = "red", size = 2) +
        theme_bw() + 
        ggtitle("Distribution Comparison - Theoretical vs Sampled (1000 Simulations)") +
        annotate("text", x = 6, y = 0.5, label = annotate4, hjust = 0) +
        annotate("text", x = 6, y = 0.45, label = annotate5, hjust = 0)
```

As I expected, they are close, albeit not perfectly aligned.

\pagebreak

# Appendices

## 100k Simulations!

Let's try it with 100k simulations, instead of 1000. Maybe the final
distribution would look closer to a normal even still!

```{r simulation2}
lambda <- 0.2 # Setting the rate for the exponential distribution
n <- 40 # Setting the number of samples to take, per average
simulations <- 100000 # Setting the number of total simulations

# Creating the overall simulations
matrix_simulations_100000 <- matrix(data = rexp(n*simulations, rate = lambda),
                             nrow = simulations)

# Creating the averages
simulation_means_100000 <- data.frame(means = apply(matrix_simulations_100000, 1, mean))
```

```{r distributioncomparison2, fig.width = 10, fig.cap = "Comparison of Sampled Distribution and Expected Theoretical Normal Distribution, 100000 simulations"}
ggplot(data = simulation_means_100000, aes(x = means)) +
        geom_histogram(aes(y = ..density..),
                       binwidth = 0.1,
                       col = "black",
                       fill = "white") + 
        geom_density(col = "blue",
                     size = 2,
                     fill = "blue",
                     alpha = 0.3) + 
        stat_function(fun = dnorm,
                      args = list(mean = 1/lambda, sd = sqrt(theoretical_variance)),
                      colour = "red", size = 2) +
        theme_bw() + 
        ggtitle("Distribution Comparison - Theoretical vs Sampled (100000 Simulations)") +
        annotate("text", x = 6, y = 0.5, label = annotate4, hjust = 0) +
        annotate("text", x = 6, y = 0.45, label = annotate5, hjust = 0)
```

Much closer!

## 100k Simulations with 100 data points per simulation!

Let's try it with 100k simulations, instead of 1000. Maybe the final
distribution would look closer to a normal even still!

```{r simulation3}
lambda <- 0.2 # Setting the rate for the exponential distribution
n <- 100 # Setting the number of samples to take, per average
simulations <- 100000 # Setting the number of total simulations

# Creating the overall simulations
matrix_simulations_100000 <- matrix(data = rexp(n*simulations, rate = lambda),
                             nrow = simulations)

# Creating the averages
simulation_means_100000 <- data.frame(means = apply(matrix_simulations_100000, 1, mean))
```

```{r distributioncomparison3, fig.width = 10, fig.cap = "Comparison of Sampled Distribution and Expected Theoretical Normal Distribution, 100000 simulations, 100 data points per simulation"}
theoretical_variance <- (1/(lambda*sqrt(n)))^2

ggplot(data = simulation_means_100000, aes(x = means)) +
        geom_histogram(aes(y = ..density..),
                       binwidth = 0.1,
                       col = "black",
                       fill = "white") + 
        geom_density(col = "blue",
                     size = 2,
                     fill = "blue",
                     alpha = 0.3) + 
        stat_function(fun = dnorm,
                      args = list(mean = 1/lambda, sd = sqrt(theoretical_variance)),
                      colour = "red", size = 2) +
        theme_bw() + 
        ggtitle("Distribution Comparison - Theoretical vs Sampled (100000 Simulations, 100pts/Simulation)") +
        annotate("text", x = 6, y = 0.5, label = annotate4, hjust = 0) +
        annotate("text", x = 6, y = 0.45, label = annotate5, hjust = 0)
```

Much closer still!


***

## Additional Resources

* [Github Repository](https://github.com/bkkesseler/DSS6-Statistical-Inference-Week4)

## Session Information

- Toshiba Chromeboook 2 (2015)
- 1.70 GHz Intel Celeron 3215U
- 4GB RAM
- RStudio Version 0.99.893

```{r sessioninfo}
sessionInfo()
```